Online Multiclass Linear Classification with Bandit Feedback

### 2019-01-09 ICML-submission/appendix-multiclass-perceptron.tex
Add algorithm for strongly separable data
Rename linear separability to weak linear separability

\begin{algorithm}[h]
\caption{\textsc{Multiclass Perceptron}
\label{algorithm:mutliclass-perceptron}}
\begin{algorithmic}[1]
{
\REQUIRE{Number of classes $K$. Number of rounds $T$. Inner product space $(V,\ip{\cdot}{\cdot})$.}
\STATE{Initialize $w_1^{(1)} = w_2^{(1)} = \dots = w_K^{(1)} = 0$}
\FOR{$t=1,2,\dots,T$}
\STATE{Observe feature vector $x_t \in V$}
\STATE{Predict $\widehat y_t = \argmax_{i \in \{1,2,\dots,K\}} \ip{w_t^{(i)}}{x_t}$}
\STATE{Observe $y_t \in \{1,2,\dots,K\}$}
\IF{$\widehat y_t \neq y_t$}
\STATE{Set $w_i^{(t+1)} = w_i^{(t)}$ for all $i \in \{1,2,\dots,K\} \setminus \{y_t, \widehat y_t\}$}
\STATE{Update $w_{y_t}^{(t+1)} = w_{y_t}^{(t)} + x_t$}
\STATE{Update $w_{\widehat y_t}^{(t+1)} = w_{\widehat y_t}^{(t)} - x_t$}
\ELSE
\STATE{Set $w_i^{(t+1)} = w_i^{(t)}$ for all $i \in \{1,2,\dots,K\}$}
\ENDIF
\ENDFOR
}
\end{algorithmic}
\end{algorithm}

\begin{theorem}[Mistake upper bound]
\label{theorem:mutliclass-perceptron-mistake-upper-bound}
Let $(V, \ip{\cdot}{\cdot})$ be an inner product space, let $K$ be a positive
integer, let $\gamma$ be a positive real number and let $R$ be a non-negative real
number. If $(x_1, y_1), (x_2, y_2), \dots, (x_T, y_T)$ is a sequence of labeled
examples in $V \times \{1,2,\dots,K\}$ that are linearly separable with margin
$\gamma$ and $\norm{x_1}, \norm{x_2}, \dots, \norm{x_T} \le R$
then \textsc{Multiclass Perceptron} algorithm makes at most $\lfloor
2(R/\gamma)^2 \rfloor$ mistakes.
\end{theorem}

\begin{proof}
Let $M = \sum_{t=1}^T \indicator{\widehat y_t \neq y_t}$ be the number of
mistakes the algorithm makes. Since the $K$-tuple $(w_1^{(t)}, w_2^{(t)}, \dots,
w_K^{(t)})$ changes only if a mistake is made, we can upper bound $\sum_{i=1}^K
\norm{w_i^{(t)}}^2$ in terms of number of mistakes.
If a mistake happens in round $t$ then
\begin{align*}
& \sum_{i=1}^K \norm{w_i^{(t+1)}}^2 \\
& = \left(\sum_{i \in \{1,2,\dots,K\} \setminus \{y_t, \widehat y_t\} } \norm{w_i^{(t)}}^2 \right) \\
& \qquad + \norm{w_{y_t}^{(t)} + x_t}^2 + \norm{w_{\widehat y_t}^{(t)} - x_t}^2 \\
& = \left(\sum_{i \in \{1,2,\dots,K\} \setminus \{y_t, \widehat y_t\} } \norm{w_i^{(t)}}^2 \right) + \norm{w_{y_t}^{(t)}}^2 + \norm{w_{\widehat y_t}^{(t)}}^2 \\
& \qquad + 2 \norm{x_t}^2 + 2 \ip{w_{y_t}^{(t)} - w_{\widehat y_t}^{(t)}}{x_t} \\
& = \left(\sum_{i=1}^K \norm{w_i^{(t)}}^2 \right) + 2 \norm{x_t}^2 + 2 \ip{w_{y_t}^{(t)} - w_{\widehat y_t}^{(t)}}{x_t} \\
& \le \left(\sum_{i=1}^K \norm{w_i^{(t)}}^2 \right) + 2 \norm{x_t}^2 \\
& \le \left(\sum_{i=1}^K \norm{w_i^{(t)}}^2 \right) + 2 R^2 \; .
\end{align*}
So each time a mistake happens, $\sum_{i=1}^K \norm{w_i^{(t)}}^2$ increases by at most $2R^2$. Thus,
$$
\sum_{i=1}^K \norm{w_i^{(T+1)}}^2 \le 2R^2 M \; .
$$
Let $w_1^*, w_2^*, \dots, w_K^* \in V$ be vectors satisfying the
\eqref{equation:linear-separability-1} and
\eqref{equation:linear-separability-2}. We lower bound $\sum_{i=1}^K \ip{w_i^*}{w_i^{(t)}}$. This quantity changes
only when a mistakes happens. If mistake happens in round $t$, we have
\begin{align*}
& \sum_{i=1}^K \ip{w_i^*}{w_i^{(t+1)}} \\
& = \left( \sum_{i \in \{1,2,\dots,K\} \setminus \{y_t, \widehat y_t\}} \ip{w_i^*}{w_i^{(t)}} \right) \\
& \qquad + \ip{w_{y_t}^*}{w_{y_t}^{(t)} + x_t} + \ip{w_{\widehat y_t}^*}{w_{\widehat y_t}^{(t)} - x_t} \\
& = \left( \sum_{i=1}^K \ip{w_i^*}{w_i^{(t)}} \right) + \ip{w_{y_t}^* - w_{\widehat y_t}^*}{x_t} \\
& \ge  \left( \sum_{i=1}^K \ip{w_i^*}{w_i^{(t)}} \right) + \gamma \; .
\end{align*}
Thus, after $M$ mistakes,
$$
\sum_{i=1}^K \ip{w_i^*}{w_i^{(T+1)}} \ge \gamma M \; .
$$
We upper bound the left hand side by using Cauchy-Schwartz inequality twice and
the condition \eqref{equation:linear-separability-1} on $w_1^*, w_2^*, \dots,
w_K^*$. We have
\begin{align*}
\sum_{i=1}^K \ip{w_i^*}{w_i^{(T+1)}}
& \le \sum_{i=1}^K \norm{w_i^*} \cdot \norm{w_i^{(T+1)}} \\
& \le \sqrt{\sum_{i=1}^K \norm{w_i^*}^2} \sqrt{\sum_{i=1}^K \norm{w_i^{(T+1)}}^2} \\
& \le \sqrt{\sum_{i=1}^K \norm{w_i^{(T+1)}}^2} \; .
\end{align*}
Combining all inequalities, we get
$$
(\gamma M)^2 \le \sum_{i=1}^K \norm{w_i^{(T+1)}}^2 \le 2R^2 M \; .
$$
We conclude that $M \le 2(R/\gamma)^2$. Since $M$ is an integer, $M \le \lfloor 2(R/\gamma)^2 \rfloor$.
\end{proof}

\begin{theorem}[Mistake lower bound]
\label{theorem:online-multiclass-classification-mistake-lower-bound}
Let $K$ be a positive integer, let $\gamma$ be a positive real number and let
$R$ be a non-negative real number. For any deterministic algorithm for the
\textsc{Online Multiclass Classification} problem there exists an
inner product space $(V, \ip{\cdot}{\cdot})$, a non-negative integer $T$ and a
sequence of labeled examples $(x_1, y_1), (x_2, y_2), \dots, (x_T, y_T)$
examples in $V \times \{1,2,\dots,K\}$ that are linearly separable with margin
$\gamma$, the norms satisfy $\norm{x_1}, \norm{x_2}, \dots, \norm{x_T} \le R$
and the algorithm makes at least $\lfloor (R/\gamma)^2 \rfloor$ mistakes.
\end{theorem}

@@

It remains to show that the examples we have constructed are linearly separable
with margin $\gamma$. To prove that we demonstrate vectors $w_1, w_2, \dots, w_K$
satisfying conditions \eqref{equation:linear-separability-1} and
\eqref{equation:linear-separability-2}. We define
$$
w_i = \frac{1}{R\sqrt{T}} \sum_{\substack{t : 1 \le t \le T \\ y_t = i}} x_t \qquad \text{for $i=1,2,\dots,K$.}
$$
Let $a_i = |\{ t ~:~ 1 \le t \le T, \ y_t = i \}|$ be the number of occurrences of label $i$.
It is easy to see that
$$
\norm{w_i}^2 = \frac{1}{R^2 T} \sum_{\substack{t : 1 \le t \le T \\ y_t = i}} \norm{x_t}^2 = \frac{a_i}{T} \qquad \text{for $i=1,2,\dots,K$.}
$$
Since $\sum_{i=1}^K a_i = T$, the condition
\eqref{equation:linear-separability-1} holds. To verify condition
\eqref{equation:linear-separability-2} consider any labeled example $(x_t,
y_t)$. Then, by definition of $w_{y_t}$ we have
\begin{align*}
\ip{w_{y_t}}{x_t}
& = \frac{1}{R\sqrt{T}} \sum_{\substack{s : 1 \le s \le T \\ y_s = y_t}} \ip{x_s}{x_t} \\
& = \frac{1}{R\sqrt{T}} \norm{x_t}^2 \\
& = \frac{R}{\sqrt{T}} \\
& \ge \frac{R}{R/\gamma} \\
& = \gamma \; .
\end{align*}
and for any $i \in \{1,2,\dots,K\} \setminus \{y_t\}$ we have
$\ip{w_i}{x_t} = 0$. Therefore, condition \eqref{equation:linear-separability-2} holds.
\end{proof}

>>>>

\begin{algorithm}[h]
\caption{\textsc{Multiclass Perceptron}
\label{algorithm:mutliclass-perceptron}}
\begin{algorithmic}[1]
{
\REQUIRE{Number of classes $K$. Number of rounds $T$. Inner product space $(V,\ip{\cdot}{\cdot})$.}
\STATE{Initialize $w_1^{(1)} = w_2^{(1)} = \dots = w_K^{(1)} = 0$}
\FOR{$t=1,2,\dots,T$}
\STATE{Observe feature vector $x_t \in V$}
\STATE{Predict $\widehat y_t = \argmax_{i \in \{1,2,\dots,K\}} \ip{w_t^{(i)}}{x_t}$}
\STATE{Observe $y_t \in \{1,2,\dots,K\}$}
\IF{$\widehat y_t \neq y_t$}
\STATE{Set $w_i^{(t+1)} = w_i^{(t)}$ \\ \qquad for all $i \in \{1,2,\dots,K\} \setminus \{y_t, \widehat y_t\}$}
\STATE{Update $w_{y_t}^{(t+1)} = w_{y_t}^{(t)} + x_t$}
\STATE{Update $w_{\widehat y_t}^{(t+1)} = w_{\widehat y_t}^{(t)} - x_t$}
\ELSE
\STATE{Set $w_i^{(t+1)} = w_i^{(t)}$ for all $i \in \{1,2,\dots,K\}$}
\ENDIF
\ENDFOR
}
\end{algorithmic}
\end{algorithm}

\begin{theorem}[Mistake upper bound]
\label{theorem:mutliclass-perceptron-mistake-upper-bound}
Let $(V, \ip{\cdot}{\cdot})$ be an inner product space, let $K$ be a positive
integer, let $\gamma$ be a positive real number and let $R$ be a non-negative real
number. If $(x_1, y_1), (x_2, y_2), \dots, (x_T, y_T)$ is a sequence of labeled
examples in $V \times \{1,2,\dots,K\}$ that are weakly linearly separable with margin
$\gamma$ and $\norm{x_1}, \norm{x_2}, \dots, \norm{x_T} \le R$
then \textsc{Multiclass Perceptron} algorithm makes at most $\lfloor
2(R/\gamma)^2 \rfloor$ mistakes.
\end{theorem}

\begin{proof}
Let $M = \sum_{t=1}^T \indicator{\widehat y_t \neq y_t}$ be the number of
mistakes the algorithm makes. Since the $K$-tuple $(w_1^{(t)}, w_2^{(t)}, \dots,
w_K^{(t)})$ changes only if a mistake is made, we can upper bound $\sum_{i=1}^K
\norm{w_i^{(t)}}^2$ in terms of number of mistakes.
If a mistake happens in round $t$ then
\begin{align*}
& \sum_{i=1}^K \norm{w_i^{(t+1)}}^2 \\
& = \left(\sum_{i \in \{1,2,\dots,K\} \setminus \{y_t, \widehat y_t\} } \norm{w_i^{(t)}}^2 \right) \\
& \qquad + \norm{w_{y_t}^{(t)} + x_t}^2 + \norm{w_{\widehat y_t}^{(t)} - x_t}^2 \\
& = \left(\sum_{i \in \{1,2,\dots,K\} \setminus \{y_t, \widehat y_t\} } \norm{w_i^{(t)}}^2 \right) + \norm{w_{y_t}^{(t)}}^2 + \norm{w_{\widehat y_t}^{(t)}}^2 \\
& \qquad + 2 \norm{x_t}^2 + 2 \ip{w_{y_t}^{(t)} - w_{\widehat y_t}^{(t)}}{x_t} \\
& = \left(\sum_{i=1}^K \norm{w_i^{(t)}}^2 \right) + 2 \norm{x_t}^2 + 2 \ip{w_{y_t}^{(t)} - w_{\widehat y_t}^{(t)}}{x_t} \\
& \le \left(\sum_{i=1}^K \norm{w_i^{(t)}}^2 \right) + 2 \norm{x_t}^2 \\
& \le \left(\sum_{i=1}^K \norm{w_i^{(t)}}^2 \right) + 2 R^2 \; .
\end{align*}
So each time a mistake happens, $\sum_{i=1}^K \norm{w_i^{(t)}}^2$ increases by at most $2R^2$. Thus,
$$
\sum_{i=1}^K \norm{w_i^{(T+1)}}^2 \le 2R^2 M \; .
$$
Let $w_1^*, w_2^*, \dots, w_K^* \in V$ be vectors satisfying the
\eqref{equation:weak-linear-separability-1} and
\eqref{equation:weak-linear-separability-2}. We lower bound $\sum_{i=1}^K \ip{w_i^*}{w_i^{(t)}}$. This quantity changes
only when a mistakes happens. If mistake happens in round $t$, we have
\begin{align*}
& \sum_{i=1}^K \ip{w_i^*}{w_i^{(t+1)}} \\
& = \left( \sum_{i \in \{1,2,\dots,K\} \setminus \{y_t, \widehat y_t\}} \ip{w_i^*}{w_i^{(t)}} \right) \\
& \qquad + \ip{w_{y_t}^*}{w_{y_t}^{(t)} + x_t} + \ip{w_{\widehat y_t}^*}{w_{\widehat y_t}^{(t)} - x_t} \\
& = \left( \sum_{i=1}^K \ip{w_i^*}{w_i^{(t)}} \right) + \ip{w_{y_t}^* - w_{\widehat y_t}^*}{x_t} \\
& \ge  \left( \sum_{i=1}^K \ip{w_i^*}{w_i^{(t)}} \right) + \gamma \; .
\end{align*}
Thus, after $M$ mistakes,
$$
\sum_{i=1}^K \ip{w_i^*}{w_i^{(T+1)}} \ge \gamma M \; .
$$
We upper bound the left hand side by using Cauchy-Schwartz inequality twice and
the condition \eqref{equation:weak-linear-separability-1} on $w_1^*, w_2^*, \dots,
w_K^*$. We have
\begin{align*}
\sum_{i=1}^K \ip{w_i^*}{w_i^{(T+1)}}
& \le \sum_{i=1}^K \norm{w_i^*} \cdot \norm{w_i^{(T+1)}} \\
& \le \sqrt{\sum_{i=1}^K \norm{w_i^*}^2} \sqrt{\sum_{i=1}^K \norm{w_i^{(T+1)}}^2} \\
& \le \sqrt{\sum_{i=1}^K \norm{w_i^{(T+1)}}^2} \; .
\end{align*}
Combining all inequalities, we get
$$
(\gamma M)^2 \le \sum_{i=1}^K \norm{w_i^{(T+1)}}^2 \le 2R^2 M \; .
$$
We conclude that $M \le 2(R/\gamma)^2$. Since $M$ is an integer, $M \le \lfloor 2(R/\gamma)^2 \rfloor$.
\end{proof}

\begin{theorem}[Mistake lower bound]
\label{theorem:online-multiclass-classification-mistake-lower-bound}
Let $K$ be a positive integer, let $\gamma$ be a positive real number and let
$R$ be a non-negative real number. For any deterministic algorithm for the
\textsc{Online Multiclass Classification} problem there exists an
inner product space $(V, \ip{\cdot}{\cdot})$, a non-negative integer $T$ and a
sequence of labeled examples $(x_1, y_1), (x_2, y_2), \dots, (x_T, y_T)$
examples in $V \times \{1,2,\dots,K\}$ that are weakly linearly separable with margin
$\gamma$, the norms satisfy $\norm{x_1}, \norm{x_2}, \dots, \norm{x_T} \le R$
and the algorithm makes at least $\lfloor (R/\gamma)^2 \rfloor$ mistakes.
\end{theorem}

@@

It remains to show that the examples we have constructed are linearly separable
with margin $\gamma$. To prove that we demonstrate vectors $w_1, w_2, \dots, w_K$
satisfying conditions \eqref{equation:weak-linear-separability-1} and
\eqref{equation:weak-linear-separability-2}. We define
$$
w_i = \frac{1}{R\sqrt{T}} \sum_{\substack{t : 1 \le t \le T \\ y_t = i}} x_t \qquad \text{for $i=1,2,\dots,K$.}
$$
Let $a_i = |\{ t ~:~ 1 \le t \le T, \ y_t = i \}|$ be the number of occurrences of label $i$.
It is easy to see that
$$
\norm{w_i}^2 = \frac{1}{R^2 T} \sum_{\substack{t : 1 \le t \le T \\ y_t = i}} \norm{x_t}^2 = \frac{a_i}{T} \qquad \text{for $i=1,2,\dots,K$.}
$$
Since $\sum_{i=1}^K a_i = T$, the condition
\eqref{equation:weak-linear-separability-1} holds. To verify condition
\eqref{equation:weak-linear-separability-2} consider any labeled example $(x_t,
y_t)$. Then, by definition of $w_{y_t}$ we have
\begin{align*}
\ip{w_{y_t}}{x_t}
& = \frac{1}{R\sqrt{T}} \sum_{\substack{s : 1 \le s \le T \\ y_s = y_t}} \ip{x_s}{x_t} \\
& = \frac{1}{R\sqrt{T}} \norm{x_t}^2 \\
& = \frac{R}{\sqrt{T}} \\
& \ge \frac{R}{R/\gamma} \\
& = \gamma \; .
\end{align*}
and for any $i \in \{1,2,\dots,K\} \setminus \{y_t\}$ we have
$\ip{w_i}{x_t} = 0$. Therefore, condition \eqref{equation:weak-linear-separability-2} holds.
\end{proof}

### 2019-01-09 ICML-submission/introduction.tex
Add algorithm for strongly separable data
Rename linear separability to weak linear separability

We study the problem of \textsc{Online Multiclass Linear Classification with
Bandit Feedback}~\citep{Kakade-Shalev-Shwartz-Tewari-2008}. The problem can be
viewed as a game between the learner and the adversary: at each timestep $t$,
the adversary chooses an example $(x_t, y_t)$ where the feature vector $x_t \in
\R^d$ is revealed to the learner and the correct label $y_t$ is kept secret by
the adversary. Upon receiving the feature vector $x_t$, the learner is asked to
make a prediction $\widehat{y}_t$. Immediately after making its prediction, the
learner receives a feedback. In contrast with the standard full information
setting (where the feedback given is the correct label $y_t$), here the feedback
is only a binary indicator of whether the prediction was correct or not. The
protocol of the problem is formally stated below.

\begin{algorithm}[h]
\caption{\textsc{Online Multiclass Classification with Bandit Feedback}
\label{algorithm:game-protocol}}
\begin{algorithmic}[1]
{
\REQUIRE{Number of classes $K$. Number of rounds $T$.}
\FOR{$t=1,2,\dots,T$}
\STATE{Observe feature vector $x_t$}
\STATE{Adversary \emph{secretly} chooses correct class label $y_t \in \{1,2,\dots,K\}$}
\chicheng{Or we simply say: Adversary chooses example $(x_t, y_t)$, where $x_t$ is reveal to the learner.}
\STATE{Predict class label $\widehat y_t \in \{1,2,\dots,K\}$}
\STATE{Observe feedback $z_t \in \{0,1\}$ where $z_t = \begin{cases}
0 & \text{if $\widehat y_t = y_t$,} \\
1 & \text{if $\widehat y_t \neq y_t$.}
\end{cases}$
}
\ENDFOR
}
\end{algorithmic}
\end{algorithm}

@@

In this paper, we focus on the special case when the examples chosen by the
adversary lie in $\R^d$ and
are \emph{linearly separable with a margin}
(see Definition~\ref{def:linear-sep} for a formal description).
%\chicheng{I wonder if we should introduce the two notions of linear separability here,
%because strong linear separability is also worth mentioning in the full-information setting.
%}
%\david{The formal definitions are in a section of their own.}

In the full-information feedback
setting, it is well known that if the examples have norm at most $R$ and are
linearly separable with a margin $\gamma$ then the \textsc{Multiclass
Perceptron} algorithm makes at most $\lfloor 2(R/\gamma)^2 \rfloor$ mistakes.
This result is very satisfying since the upper bound on the number of mistakes
is information-theoretically optimal and at the same time the \textsc{Multiclass
Perceptron} algorithm has low time and memory complexity.

The bandit feedback setting, however, is much more challenging. To understand
the differences, we dissect the notion of linear separability in the multiclass
setting. Besides the standard notion of linear separability, we also consider
\emph{strong linear separability} (see Definition~\ref{def:strong-linear-sep} for a formal description). For binary classification, these two
definitions are identical.
However, they differ for multiclass classification with $K \geq 3$.
For multiclass classification with $K$ classes, the standard notion of linear
separability means that the examples from each class lie in an intersection of
$K-1$ halfspaces and the examples outside of the class lie in the complement of
the intersection of the halfspaces. Strong linear separability means that
examples from each class are separated from the remaining examples by a
\emph{single} hyperplane. We state the definitions precisely in
Section~\ref{section:notions-of-linear-separability}.

For the case when the examples are strongly linearly separable, we design a
simple and efficient algorithm with an expected number of mistakes at most $(K-1)
\lfloor (R/\gamma)^2 \rfloor$. The algorithm can be viewed as running $K$ copies
of the \textsc{Binary Perceptron} algorithm, one copy for each class.
Intuitively speaking, the factor $K-1$ is the price we pay for the bandit
feedback, or more precisely, the lack of full-information feedback. We also
prove that any (possibly randomized) algorithm makes $\Omega(K (R/\gamma)^2))$
mistakes in the worst case.

>>>>

We study the problem of \textsc{Online Multiclass Linear Classification with
Bandit Feedback}~\citep{Kakade-Shalev-Shwartz-Tewari-2008}. The problem can be
viewed as a repeated game between the learner and the adversary. At each timestep $t$,
the adversary chooses a labeled example $(x_t, y_t)$ where the feature vector, $x_t \in
\R^d$ is revealed to the learner and the correct label $y_t$ is kept secret by
the adversary. Upon receiving the feature vector $x_t$, the learner is asked to
make a prediction $\widehat{y}_t$. Immediately after making its prediction, the
learner receives a feedback. In contrast with the standard full information
setting (where the feedback given is the correct label $y_t$), here the feedback
is only a binary indicator of whether the prediction was correct or not. The
protocol of the problem is formally stated below.

\begin{algorithm}[h]
\caption{\textsc{Online Multiclass Classification with Bandit Feedback}
\label{algorithm:game-protocol}}
\begin{algorithmic}[1]
{
\REQUIRE{Number of classes $K$.}
\REQUIRE{Number of rounds $T$.}
\FOR{$t=1,2,\dots,T$}
\STATE{Observe feature vector $x_t$}
\STATE{Adversary \emph{secretly} chooses correct class label $y_t \in \{1,2,\dots,K\}$}
\chicheng{Or we simply say: Adversary chooses example $(x_t, y_t)$, where $x_t$ is reveal to the learner.}
\STATE{Predict class label $\widehat y_t \in \{1,2,\dots,K\}$}
\STATE{Observe feedback $z_t \in \{0,1\}$ \\ \qquad where $z_t = \begin{cases}
0 & \text{if $\widehat y_t = y_t$,} \\
1 & \text{if $\widehat y_t \neq y_t$.}
\end{cases}$
}
\ENDFOR
}
\end{algorithmic}
\end{algorithm}

@@

In this paper, we focus on the special case when the examples chosen by the
adversary lie in $\R^d$ and are linearly separable with a margin. We introduce
two notions of linear separability, \emph{weak} and \emph{strong}. These are
formally stated in
Definitions~\ref{definition:weak-linear-separability}~and~\ref{definition:strong-linear-separability}.
The standard notion of multiclass linear separability corresponds to the weak
linear separability. For binary classification, these two notions are
identical. However, they differ for multiclass classification with $K \geq 3$.
For multiclass classification with $K$ classes, the standard notion of linear
separability means that the examples from each class lie in an intersection of
$K-1$ halfspaces and the examples outside of the class lie in the complement of
the intersection of the halfspaces. Strong linear separability means that
examples from each class are separated from the remaining examples by a
\emph{single} hyperplane.

In the full-information feedback setting, it is well known that if the examples
have norm at most $R$ and are weakly linearly separable with a margin $\gamma$
then the \textsc{Multiclass Perceptron} algorithm makes at most $\lfloor
2(R/\gamma)^2 \rfloor$ mistakes. This result is very satisfying since the upper
bound on the number of mistakes is information-theoretically optimal and at the
same time the \textsc{Multiclass Perceptron} algorithm has low time and memory
complexity.

The bandit feedback setting, however, is much more challenging. For the case
when the examples are strongly linearly separable, we design a simple and
efficient algorithm with an expected number of mistakes at most $(K-1) \lfloor
(R/\gamma)^2 \rfloor$. The algorithm can be viewed as running $K$ copies of the
\textsc{Binary Perceptron} algorithm, one copy for each class. Intuitively
speaking, the factor $K-1$ is the price we pay for the bandit feedback, or more
precisely, the lack of full-information feedback. We also prove that any
(possibly randomized) algorithm makes $\Omega(K (R/\gamma)^2))$ mistakes in the
worst case.

### 2019-01-09 ICML-submission/notions-of-linear-separability.tex
Rename linear separability to weak linear separability

\begin{definition}[Linear separability]
\label{definition:linear-separability}
Let $(V,\ip{\cdot}{\cdot})$ be an inner product space, $K$ be a positive
integer, and $\gamma$ be a positive real number. Let $(x_1, y_1), (x_2,
y_2), \dots, (x_T, y_T)$ be labeled examples in $V \times \{1,2,\dots,K\}$.

The examples are said to be \emph{linearly separable with a
margin $\gamma$} if there exist vectors $w_1, w_2, \dots, w_K \in V$ such
that
\begin{align}
\label{equation:linear-separability-1}
& \sum_{i=1}^K \norm{w_i}^2 \le 1 \; , \\
\label{equation:linear-separability-2}
& \begin{gathered}
\forall t \in \{1,2,\dots,T\} \quad \forall i \in \{1,2,\dots, K\} \setminus \{y_t\} \\
\ip{x_t}{w_{y_t}} \ge \ip{x_t}{w_i} + \gamma \; .
\end{gathered}
\end{align}
\end{definition}

@@

The notion of linear separability with a margin is standard. It is used in the
full-information setting to upper bound the number of mistakes of the
\textsc{Multiclass Perceptron} algorithm. For completeness, we state the
algorithm in Appendix~\ref{section:multiclass-perceptron-proofs}. It is a
folklore result that if a set of labeled examples is separable with a margin
$\gamma$ and the norm of the examples is bounded by $R$, then \textsc{Multiclass
Perceptron} algorithm makes at most $\left\lfloor 2(R/\gamma)^2 \right \rfloor$
mistakes. Another folklore result is that \textsc{Multiclass Perceptron} is
essentially optimal in the sense that any deterministic algorithm must make
$\left\lfloor (R/\gamma)^2 \right \rfloor$ mistakes in the worst case. For
completeness, we give proofs of both of these results in
Appendix~\ref{section:multiclass-perceptron-proofs}.

@@

It is easy to see that if a set of labeled examples are strongly linearly
separable with margin $\gamma$, then they are linearly separable with margin
$2\gamma$. Indeed, if $w_1, w_2, \dots, w_K \in V$
satisfy \eqref{equation:strong-linear-separability-1},
\eqref{equation:strong-linear-separability-2},
\eqref{equation:strong-linear-separability-3} then they satisfy
\eqref{equation:linear-separability-1} and
\eqref{equation:linear-separability-2} with margin $\gamma' = 2\gamma$.

Similarly, if $K=2$ and labeled examples are linearly separable with a margin
$\gamma$ then they are strongly linearly separable with margin $\gamma/2$.
Indeed, if $w_1, w_2$ satisfy \eqref{equation:linear-separability-1} and
\eqref{equation:linear-separability-2} then $w_1' = \frac{w_1 - w_2}{2}$, $w_2' =
\frac{w_2 - w_1}{2}$ and $\gamma'=\gamma/2$ satisfy
\eqref{equation:strong-linear-separability-1},
\eqref{equation:strong-linear-separability-2},
\eqref{equation:strong-linear-separability-3}. The last two conditions are easy
to see. The first condition follows from $\norm{w_i'}^2 \le (\frac{1}{2}
\norm{w_1} + \frac{1}{2} \norm{w_2})^2 \le \frac{1}{2}\norm{w_1}^2 +
\frac{1}{2}\norm{w_2}^2 \le \frac{1}{2}$ for $i=1,2$.

@@

However, for any $K \ge 3$ and any inner product space of dimension at least
$2$, there exists a set of labeled examples that is linearly separable with a
positive margin $\gamma$ but it is not strongly linearly separable with any
positive margin $\gamma$. An example of such set of labeled examples is shown in
Figure~\ref{figure:linearly-separable-examples-with-margin}.

\begin{figure}
\begin{center}
\input{figures/linearly-separable-examples-with-margin}
\end{center}
\caption[]{The figure shows labeled examples in $\R^2$. The examples belong to
$K=3$ classes colored white, gray and black respectively. Each class lies in a
$120^\circ$ wedge. In other words, each class lies in an intersection of two
halfspaces.

While the examples are linearly separable with a positive margin $\gamma$, they
are \emph{not} strongly linearly separable with any positive margin $\gamma$.
}
\label{figure:linearly-separable-examples-with-margin}
\end{figure}

>>>>

\begin{definition}[Weak linear separability]
\label{definition:weak-linear-separability}
Let $(V,\ip{\cdot}{\cdot})$ be an inner product space, $K$ be a positive
integer, and $\gamma$ be a positive real number. Let $(x_1, y_1), (x_2,
y_2), \dots, (x_T, y_T)$ be labeled examples in $V \times \{1,2,\dots,K\}$.

The examples are said to be \emph{weakly linearly separable with a
margin $\gamma$} if there exist vectors $w_1, w_2, \dots, w_K \in V$ such
that
\begin{align}
\label{equation:weak-linear-separability-1}
& \sum_{i=1}^K \norm{w_i}^2 \le 1 \; , \\
\label{equation:weak-linear-separability-2}
& \begin{gathered}
\forall t \in \{1,2,\dots,T\} \quad \forall i \in \{1,2,\dots, K\} \setminus \{y_t\} \\
\ip{x_t}{w_{y_t}} \ge \ip{x_t}{w_i} + \gamma \; .
\end{gathered}
\end{align}
\end{definition}

@@

The notion of weak linear separability with a margin is standard. It is used in the
full-information setting to upper bound the number of mistakes of the
\textsc{Multiclass Perceptron} algorithm. It is a
folklore result that if a set of labeled examples is separable with a margin
$\gamma$ and the norm of the examples is bounded by $R$, then \textsc{Multiclass
Perceptron} algorithm makes at most $\left\lfloor 2(R/\gamma)^2 \right \rfloor$
mistakes. Another folklore result is that \textsc{Multiclass Perceptron} is
essentially optimal in the sense that any deterministic algorithm must make
$\left\lfloor (R/\gamma)^2 \right \rfloor$ mistakes in the worst case. For
completeness, we state the \textsc{Multiclass Perceptron} algorithm
and give proofs of both of these results in
Appendix~\ref{section:multiclass-perceptron-proofs}.

@@

It is easy to see that if a set of labeled examples are strongly linearly
separable with margin $\gamma$, then they are weakly linearly separable with margin
$2\gamma$. Indeed, if $w_1, w_2, \dots, w_K \in V$
satisfy \eqref{equation:strong-linear-separability-1},
\eqref{equation:strong-linear-separability-2},
\eqref{equation:strong-linear-separability-3} then they satisfy
\eqref{equation:weak-linear-separability-1} and
\eqref{equation:weak-linear-separability-2} with margin $\gamma' = 2\gamma$.

Similarly, if $K=2$ and labeled examples are weakly linearly separable with a margin
$\gamma$ then they are strongly linearly separable with margin $\gamma/2$.
Indeed, if $w_1, w_2$ satisfy \eqref{equation:weak-linear-separability-1} and
\eqref{equation:weak-linear-separability-2} then $w_1' = \frac{w_1 - w_2}{2}$, $w_2' =
\frac{w_2 - w_1}{2}$ and $\gamma'=\gamma/2$ satisfy
\eqref{equation:strong-linear-separability-1},
\eqref{equation:strong-linear-separability-2},
\eqref{equation:strong-linear-separability-3}. The last two conditions are easy
to see. The first condition follows from $\norm{w_i'}^2 \le (\frac{1}{2}
\norm{w_1} + \frac{1}{2} \norm{w_2})^2 \le \frac{1}{2}\norm{w_1}^2 +
\frac{1}{2}\norm{w_2}^2 \le \frac{1}{2}$ for $i=1,2$.

@@

However, for any $K \ge 3$ and any inner product space of dimension at least
$2$, there exists a set of labeled examples that is weakly linearly separable with a
positive margin $\gamma$ but it is not strongly linearly separable with any
positive margin $\gamma$. An example of such set of labeled examples is shown in
Figure~\ref{figure:weakly-linearly-separable-examples-with-margin}.

\begin{figure}
\begin{center}
\input{figures/weakly-linearly-separable-examples-with-margin}
\end{center}
\caption[]{The figure shows labeled examples in $\R^2$. The examples belong to
$K=3$ classes colored white, gray and black respectively. Each class lies in a
$120^\circ$ wedge. In other words, each class lies in an intersection of two
halfspaces.

While the examples are weakly linearly separable with a positive margin $\gamma$, they
are \emph{not} strongly linearly separable with any positive margin $\gamma$.
}
\label{figure:weakly-linearly-separable-examples-with-margin}
\end{figure}
